# Bioconductor

:::: {.learn .panel}

::: content
- Learn about Bioconductor, a repository for R packages
- Learn about the `BiocManager` package and how to use it to install R packages hosted by Bioconductor
:::

::::

## What is Bioconductor?

[Bioconductor](https://www.bioconductor.org/) is a repository of R packages. All Bioconductor packages are regularly built and tested to ensure that they are functioning correctly and meet certain standards. 

For every new release of R a snapshot of the entire Bioconductor repository is stored and retained in order to make analysis reproducible at the time it was run.

It is essential therefore that you keep up-to-date with the R version, otherwise you will not be able to use the latest version of packages in the Bioconductor repository.

The `BiocManager` package ([link](https://www.rdocumentation.org/packages/BiocManager)) contains a small number of functions to install R packages from Bioconductor (and other repositories such as CRAN) and check that the versions are compatible with the version of R you are using.

## Installing Bioconductor packages

Using `BiocManager` to install the packages needed for this course will ensure that you are using the most recent release, and can also install the appropriate versions of all needed dependencies. 


::: {.task .panel}

::: title
Install packages
:::

:::: content
To install the packages for this course copy and paste the code below into the command line and press Enter:

```{r, eval=FALSE}
BiocManager::install(
    c('struct','structToolbox','metabolomicsWorkbenchR'),
    dependencies='suggests')
```

The `dependencies` input ensures that we install all suggested packages required to use the ones we are installing. There will be quite a few of them!

You can check that the packages are installed and working by running the following code:

```{r,message=FALSE,warning=FALSE}
library('struct')
library('structToolbox')
library('metabolomicsWorkbenchR')
```

You might receive some (sometimes cryptic!) messages after activating the packages, but as long as you receive no error messages, then you are ready to go.
::::
:::

## Additional packages
In addition to the `struct` and `structToolbox` packages it is useful to load the `ggplot2` package. It will allow you to modify e.g. chart titles quickly and easily. It should have been installed already as a dependency of `structToolbox`, so you can just activate it for the current session.

In addition you will need the following packages:

```{r,message=FALSE,warning=FALSE}
library(ggplot2)
```

In addition, you will also need make sure the following R packages are installed:

 - `dplyr`
 - `xcms`
 - `MSnbase`
 - `openxlsx`
 - `RColorBrewer`
 - `ggplot2`

::: {.critical .panel}

:::: content
We _strongly_ recommend you continue to use `BiocManager:install` rather than `install.packages` to install additional packages if possible, even if they are not a Bioconductor package (e.g. CRAN packages). This will ensure that a version compatible with your R and Bioconductor versions is used.
::::

:::
