--- 
title: "Introduction to Untargeted Metabolomics Data Analaysis in R using struct and structToolbox packages"
author: "Gavin Rhys Lloyd, Ossama Edbali and Ralf Weber"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
# url: your book url like https://bookdown.org/yihui/bookdown
# cover-image: bmtc_logo.svg
description: |
  This is a minimal example of using the bookdown package to write a book.
  set in the _output.yml file.
  The HTML output format for this example is bookdown::gitbook,
link-citations: yes
github-repo: rstudio/bookdown-demo
output:
    bookdown::gitbook:
        includes: 
            after_body: content/bmtc.html
        toc_depth: 2
    bookdown::word_document2:
        toc: true
        toc_depth: 2
---

```{r, echo=FALSE}
library(fontawesome)
library(struct)
library(structToolbox)
library(ggplot2)

knitr::write_bib(file = 'packages.bib')
knitr::opts_chunk$set(class.source = "rcode",
                      class.output = 'console',    
                      comment = "",
                      fig.align='center',
                      fig.width = 4.5, 
                      fig.height = 5)

```

# Preface {-}

This material introduces the use of `struct` and `structToolbox` packages to analyse 
untargeted LCMS metabolomics data sets. It is primarily intended as 
required pre-course training material for attendees of the Birmingham 
Metabolomics Training Centre (BMTC), but can also be used as a standalone lesson. 
The lesson includes:

- Instructions on how to get started
- An introduction to the R-package struct and its templates
- Downloading data from Metabolomics Workbench using R-package metabolomicsWorkBenchR
- Building simple data processing workflows using the pre-built templates in structToolbox

If you have any questions, contact us by [email](mailto:bmtc@bham.ac.uk).

## Prerequisites {-}

For this course we assume that you have basic programming skills in R. 
It is not the purpose of this course to teach you R programming. If you are new to R, or need a refresher, we recommend any of the following data carpentry courses:

- [Data Analysis and Visualisation in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/01-intro-to-r.html)
- [Introduction to R](https://librarycarpentry.org/lc-r/)
- [R for Reprodicible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder)
- [Programming with R](https://swcarpentry.github.io/r-novice-inflammation)

## Using this document {-}

The material for this course has been divided into modules. You can access the modules using the navigation menu on the left hand side. To view the next page of a module you can use the left and right arrows (they look a bit like this `<` and `>`). Depending on your screen size they will either appear at the bottom of each page, or on the left and right hand sides.

::: {.learn .panel}

:::: content
At the beginning of each module a set of learning objectives is provided in a blue box like this one. 
::::

:::

:::: {.note .panel}

::: content
Notes and useful additional information will be included green boxes. 
:::

::::

::: {.critical .panel}

:::: content
Critical information may be provided in red boxes. 
::::

:::

::::: {.task .panel}

::: title
Exercises
:::

::: content
Exercises for you to complete are included in yellow boxes. Exercises will sometimes include drop-down tasks, hints and solutions. 
:::

::: {.expandable .task}
Tasks
:::

::: {.content .collapsed}
Tasks will appear like this.
:::


::: {.expandable .hint}
Hints
:::

::: {.content .collapsed}
Hints will appear like this.
:::

::: {.expandable .answer}
Solution
:::

::: {.content .collapsed}
Solutions will appear like this.
:::

:::::

R code and console output will be included in code blocks with an `<R>` symbol. You can copy/paste the code directly into R by hovering over the top right corner and clicking the button that appears.

```{r}
# this code adds two numbers and stores the result
a = 1
b = 2
c = a + b
```

Any code outputs that would be printed to console will be displayed in code blocks with a an `<output>` symbol.

```{r,echo=FALSE}
print('like this')
```
